---
title: "CHLA9_vs_CHLA10_scRNA"
author: "Thejes Nair"
date: "2025-12-18"
output: html_document
---


```{r}
library(Seurat)
library(tidyverse)
library(clusterProfiler)
library(msigdbr)
library(enrichplot)
library(org.Hs.eg.db)
library(dplyr)
library(tibble)
library(purrr)
library(ggplot2)
library(svglite)
```

*Narrowed focus on CHLA9, CHLA10 from previous scRNA Seurat work*

*CHLA9 CHLA10 UMAP*

```{r}
# read in integrated rds obj
seu_rna = readRDS("/Users/thejesnair/Desktop/KCGIP/EwS/seurat_integration_finalAnnotations.rds")
```

```{r}
# metadata cols
colnames(seu_rna@meta.data) # cell_line
```

```{r}
# subset
chla = subset(seu_rna, cell_line %in% c("CHLA9", "CHLA10"))

table(chla$cell_line)
```

```{r, fig.width = 15}
DimPlot(chla, reduction = "umap", group.by = "cell_line")
#DimPlot(chla, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
DimPlot(chla, reduction = "umap", group.by = "final_annotation", label = TRUE,
        label.size=6)
```


Evidence that CHLA10 is modestly enriched for certain rare states (clusters 3/5/8), which may reflect treatment-associated shifts, but these are low-abundance populations and need confirmation.


```{r}
# CHLA cluster distribution
tab_counts = table(chla$final_annotation, chla$cell_line)
tab_counts

# summarize by final annotation states not cluster IDs
tab_props = prop.table(tab_counts, margin = 2)
round(tab_props, 3)


#tab_state = table(chla$final_annotation, chla$cell_line)
#round(prop.table(tab_state, margin=2), 3)

df_props = as.data.frame(tab_props)
colnames(df_props) <- c("final_annotation", "cell_line", "prop")

ggplot(df_props, aes(x = cell_line, y = prop, fill = final_annotation)) +
  geom_col() +
  ylab("Proportion of cells") +
  xlab("") +
  theme_classic(base_size=15) + 
  scale_fill_manual(values = c(
  "Stress-like" = "#D55E00",
  "Mesenchymal-like" = "#E69F00",
  "Mesenchymal-like (highly proliferative)" = "#009E73",
  "Neural-like" = "#0072B2",
  "Mixed neural/mesenchymal" = "#999999",
  "Highly proliferative" = "#CC79A7"
))

ggsave("CHLA9_vs_CHLA10_transcriptional_programs.png", units = "in", width=10, height=8, dpi=900)
ggsave("CHLA9_vs_CHLA10_transcriptional_programs.svg", width=10, height=8, dpi=900)
```

CHLA9 and CHLA10 occupy a shared transcriptional state space with no evidence for the emergence of novel cell states post-treatment.

Differences between the two cell lines are driven primarily by shifts in the relative abundance of existing transcriptional states, rather than strong within-state heterogeneity or subclustering.

The most abundant and well-powered states (Stress-like and Mesenchymal-like) were selected for within-state differential expression analysis to assess treatment-associated transcriptional changes while avoiding confounding from cell-state composition.

Additional states (Neural-like and Highly proliferative) show very modest enrichment in CHLA10 relative to CHLA9, but remain low-abundance populations. Therefore, they are interpreted as supportive evidence of treatment associated shifts in program usage, rather than primary drivers of transcriptional differences.

Across all states, substantial overlap in UMAP space indicates continuity and plasticity of transcriptional programs, consistent with state transitions rather than discrete, treatment-induced subpopulations.


*Treatment is associated with redistribution of conserved transcriptional states, with only subtle shifts in program usage within rare populations. Therefore, downstream analyses focus on within-state differential expression in well-powered states.*

```{r}
DimPlot(chla, reduction="umap", group.by="final_annotation",
        split.by="cell_line", label=TRUE, repel=TRUE)
```

```{r}
#exists("seurat_integration")
seu_rna
```

```{r}
seu_rna@assays
seu_rna@reductions
```

*Find markers*

Previously used FindAllMarkers() to find DEG between all cell lines.
Since we're focusing on 2 cell lines now, use **FindMarkers**

"FindMarkers - Finds markers (differentially expressed genes) for identity classes"

"FindAllMarkers - Finds markers (differentially expressed genes) for each of the identity classes in a dataset"

https://www.biostars.org/p/409790/

Focusing on stress-like and mesenchymal-like states

```{r}
# subset the diff transcriptional state programs
sub_stress = subset(chla, subset=final_annotation =="Stress-like")
sub_mes = subset(chla, subset=final_annotation == "Mesenchymal-like")

# set idents to cell lines
Idents(sub_stress) = "cell_line"
table(sub_stress$cell_line) # cell counts for cell line cluster

# find all markers
de_stress = FindMarkers(
  sub_stress,
  ident.1 = "CHLA10",
  ident.2 = "CHLA9",
  logfc.threshold = 0.25,
  min.pct = 0.1
)

# REPEAT for mes state
Idents(sub_mes) = "cell_line"
table(sub_mes$cell_line)

de_mes = FindMarkers(
  sub_mes,
  ident.1="CHLA10",
  ident.2="CHLA9",
  logfc.threshold=0.25,
  min.pct=0.1
)
```

```{r}
de_stress
de_mes

# sort by adj p-val
de_stress2 = de_stress %>% 
  rownames_to_column("gene") %>% 
  arrange(p_val_adj)

de_mes2 = de_mes %>% 
  rownames_to_column("gene") %>% 
  arrange(p_val_adj)

# sanity check
sum(de_stress2$p_val_adj <= 0.05, na.rm = TRUE) # gene count
sum(de_mes2$p_val_adj <= 0.05, na.rm = TRUE)
```

```{r}
# table of genes
sig_stress = de_stress2 %>% 
  filter(p_val_adj <=0.05, abs(avg_log2FC) >= 0.5) %>% 
  select(gene, p_val_adj, avg_log2FC, pct.1, pct.2, p_val)

sig_mes = de_mes2 %>% 
  filter(p_val_adj <=0.05, abs(avg_log2FC) >= 0.5) %>% 
  select(gene, p_val_adj, avg_log2FC, pct.1, pct.2, p_val)

nrow(sig_stress); nrow(sig_mes)
```

Reminder:
pct.1, pct.2
fraction of cells in x cell line/state that expresses this gene

so
pct.1 = 0.300
30% of cells in CHLA10 express this gene 
pct.2 = 0.00
0% of cells in CHLA9 express this gene

```{r}
# top 50 genes
top_stress = sig_stress %>% 
  slice_max(abs(avg_log2FC), n=50)

top_mes = sig_mes %>% 
  slice_max(abs(avg_log2FC), n=50)

top_stress; top_mes
```
```{r}
# save all markers
write.csv(de_stress2, "DEG_CHLA9_vs_CHLA10_StressLike_all.csv", row.names=FALSE)
write.csv(de_mes2, "DEG_CHLA9_vs_CHLA10_MesenchymalLike_all.csv", row.names=FALSE)

# save TOP degs
write.csv(top_stress, "top50deg_CHLA9_vs_CHLA10_StressLike.csv", row.names=FALSE)
write.csv(top_mes, "top50deg_CHLA9_vs_CHLA10_MesenchymalLike.csv", row.names=FALSE)
```

*Cell cycle comparison*
Similar medians b/w two cell lines, no dramatic shift
However, CHLA10 has a wider upper tail in both S and G2M: more CHLA10 cells have a higher proliferation score

Heterogeneity in terms of proliferation intensity(?) but not a widespread change in cell cycle state b/w the two


Proportion table:
CHLA10, fewer cells in G1, more cells in G2M/S relative to CHLA9:
supports mild proliferative shift POST-treatment, but not a large shift
```{r}
VlnPlot(chla, features = c("S.Score","G2M.Score"), group.by = "cell_line", pt.size = 0)
prop.table(table(chla$Phase, chla$cell_line), margin = 2)
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

