---
title: "CHLA9_vs_CHLA10_scATAC"
author: "Thejes Nair"
date: "2025-12-27"
output: html_document
---

```{r}
library(Seurat)
library(Signac)
library(dplyr)
library(tibble)
library(GenomicRanges)
library(ggplot2)
library(biovizBase)
library(EnsDb.Hsapiens.v86)
library(ggrepel)
```

Differential chromatin accessibility between CHLA9 and CHLA10

```{r}
# clear memory, local RStudio reaches mem max bc of prev loaded .rds obj
rm(seu)
gc()
# read in integrated rds obj
seu_atac = readRDS("/Users/thejesnair/Desktop/KCGIP/EwS/seurat_integration_finalAnnotations.rds")

DefaultAssay(seu_atac) = "ATAC"
```

```{r}
# sanity check
Assays(seu_atac)
DefaultAssay(seu_atac)

# cell lines
table(seu_atac$cell_line)

# subset cell lines
seu_chla = subset(seu_atac, subset=cell_line %in% c("CHLA9", "CHLA10"))
table(seu_chla$cell_line)

# remove seu_atac from mem to free up space
rm(seu_atac); gc()
```

logistic regression differential accessibility (DA), control for fragments per cell (covariate: variable used in statistical modeling to account for factors that may influence chromatin accessibility, but not primary experimental variable being studied)

nFeature_ATAC: number of peaks detected per cell, proxy for ATAC sequencing depth, covariate

nCount_ATAC: total fragment counts per cell, sequencing-depth related as well

```{r}
colnames(seu_chla@meta.data) # look for ATAC covariate
# listed as nFeature_ATAC
# listed as nCount_ATAC

# summary of nFeature_ATAC
summary(seu_chla@nFeature_ATAC)
```

```{r}
VlnPlot(seu_chla, features = c("nCount_ATAC", "nFeature_ATAC", "FRIP", "TSSEnrichment"),
        group.by = "cell_line", pt.size = 0, ncol = 2)
```
*DIFFERENTIAL ACCESSIBILITY*
```{r}
lr_da_peaks = FindMarkers(
  object=seu_chla,
  ident.1 = "CHLA9",
  ident.2 = "CHLA10",
  test.use = "LR",
  latent.vars = "nCount_ATAC",
  min.pct=0.1
)
```

```{r}
lr_da_peaks

# reformat
lr_da_peaks2 = lr_da_peaks %>% 
  rownames_to_column("peak") %>% 
  arrange(p_val_adj) %>% 
  dplyr::select(peak, p_val_adj, avg_log2FC, pct.1, pct.2, p_val)

lr_da_peaks2

# save file
write.csv(lr_da_peaks2, "DA_peaks_CHLA9_vs_CHLA10_scATAC.csv", row.names=FALSE)
```

```{r}
# how many peaks significant
sum(lr_da_peaks2$p_val_adj <= 0.05)
head(lr_da_peaks2, 10)

```

```{r}
# volcano plot
lr_da_peaks2 %>%
  mutate(direction = case_when(
    p_val_adj <= 0.05 & avg_log2FC >= 0.25 ~ "More open in CHLA9",
    p_val_adj <= 0.05 & avg_log2FC <= -0.25 ~ "More open in CHLA10",
    TRUE ~ "Not sig"
  )) %>%
  ggplot(aes(x = avg_log2FC, y = -log10(p_val_adj), color = direction)) +
  geom_point(alpha = 0.6, size = 0.8) +
  theme_classic(base_size=15) +
  xlab("log2FC accessibility (CHLA9 − CHLA10)") +
  ylab("-log10(adj p-value)") +
  ggtitle("Differential accessibility (ATAC): CHLA9 vs CHLA10")

ggsave("DA_ATAC_CHLA9_vs_CHLA10.svg", width=10, height=8, dpi=900)
```
Volcano plot:
x-axis: avg_log2FC (accessibility)
  positive values (right), peaks mean more accessibility in CHLA9 (pre-treatment)
      condition one is "up" relative to condition two "down"
  negative values (left), peaks mean more accessibility in CHLA10 (post-treatment)
      condition one is "down" relative to condition two "up"
      
y-axis:
  higher, more statistically significant
  

areas to pay attention to:
  top-left (more open in CHLA10) <- good area to focus on??? post-treatment regulatory changes
  top-right quadrants (more open in CHLA9)
    

```{r}

# verifying that genescorematrix is part of atac and not rna b/c i keep reading conflicting things online
class(seu_chla[["ATAC"]])
seu_chla[["GeneScoreMatrix"]]@counts[1:5, 1:5]
summary(seu_chla[["GeneScoreMatrix"]]@counts@x)
summary(seu_chla[["GeneExpressionMatrix"]]@counts@x)
```

```{r}
# verifying
FeatureScatter(
  seu_chla,
  feature1 = "nCount_ATAC",
  feature2 = "nCount_GeneScoreMatrix"
) # yes it is tracking ATAC data

FeatureScatter(
  seu_chla,
  feature1 = "nCount_GeneExpressionMatrix",
  feature2 = "nCount_GeneScoreMatrix"
)
```

```{r}
# select sig/biological meaningful DA peaks
sig_peaks = lr_da_peaks2 %>% 
  filter(p_val_adj) <=0.05, abs(avg_log2FC >=0.5))
```

ATAC is an Assay not a ChromatinAssay, so it was not created using CreateChromatinAssay()
What we have is peak x cell matrix store as a plain Seurat Assay

```{r}
class(seu_chla[["ATAC"]])
DefaultAssay(seu_chla) = "ATAC"
head(rownames(seu_chla))
#Annotation(seu_chla[["ATAC]]) # won't work
```

So then what about MotifMatrix?
Most likely a motif activity/deviation matrix (TF x cell)
motif-level (TFs), already aggregated, not peak-based, stored as plain assay
rownames look like peaks chr1-...
```{r}
class(seu_chla[["MotifMatrix"]])
DefaultAssay(seu_chla) = "MotifMatrix"
head(rownames(seu_chla)) # yes, TFA names
```
follow this vignette
https://stuartlab.org/signac/articles/pbmc_vignette.html


```{r}
# filter significant peaks
sig_da = lr_da_peaks2 %>% 
  filter(p_val_adj <=0.05, abs(avg_log2FC) >= 0.25) # kept at 0.25 for sensitivity

nrow(sig_da)

# convert peak strings -> GRanges
peak_gr = StringToGRanges(sig_da$peak, sep = c("-", "-"))
peak_gr
```
quick pivot, want to check gene naming conventions 
```{r}
head(rownames(seu_chla[["GeneExpressionMatrix"]], 20))
sum(grepl("^ENSG", rownames(seu_chla[["GeneExpressionMatrix"]])))

peaks <- rownames(seu_chla[["ATAC"]])
max(as.numeric(sub(".*-", "", peaks))) # atac peak coordinates to determine if hg19 or h38
# going with hg38 because closest to that so going to use EnsDb
# wanted to verify bc saw either EnsDb or TxDb as options
tail(sort(peaks[grepl("^chr1-", peaks)]), 5)
```
hg19 chr1 ≈ 249,250,621
hg38 chr1 ≈ 248,956,422


https://rdrr.io/bioc/ensembldb/f/vignettes/coordinate-mapping.Rmd


```{r}
DefaultAssay(seu_chla) <- "MotifMatrix"
dim(seu_chla) # features x cells
```

back to peak-> gene mapping
```{r}
# change default assay back to ATAC
DefaultAssay(seu_chla) = "ATAC"

# get genes
genes = GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)

# peaks are UCSC style (chr1), force genes to UCSC
# ensembl/ncbi style is 1-X,Y,MT
seqlevelsStyle(genes) = "UCSC"
genome(genes) = "hg38"
```

```{r}
# check to make sure it worked
head(seqlevels(genes))
head(seqlevels(peak_gr))
length(intersect(seqlevels(peak_gr), seqlevels(genes)))
```

peaks = genomic intervals
genes = genomic intervals

need to build spatial relationship b/w peaks and genes
StringToGRanges: parses strings (ex: chr1-59814420-59814920), splits chr, start, end
turns into GRanges object (which is the standard genomic interval container in Bioconductor)

so then each peak is a proper genomic interval, depicts the exact region on the chr
```{r}
# make sure peak_gr built from DA peak list and named
peak_gr = StringToGRanges(sig_da$peak, sep=c("-", "-")) 
# converts peak coordinate strings to genomic intervals so can be spatially compared to genes

names(peak_gr) = sig_da$peak # make sure rows are properly named to preserve peak identity for index-based operations

```

keepSeqlevels: basically, drop chr that arent in the shared list
coarse: drop chromosomes and drop ranges that are on them, so only keeps chrs in common,
discards peaks/genes on non-common chr

default is error: would throw an error and stop if anything is dropped
```{r}
# reduce to shared chromosomes
common = intersect(seqlevels(peak_gr), seqlevels(genes)) # only chr that appear in BOTH peaks and genes are kept
peak_gr = keepSeqlevels(peak_gr, common, pruning.mode = "coarse") # any genomic range that lives on a chr NOT in common is dropped entirely
# chr not shared? dropped
# oeak/gene on that chr, dropped
# chr shared? all ranges are kept (for now)

genes = keepSeqlevels(genes, common, pruning.mode="coarse")
```

```{r}
# MAP PEAK -> NEAREST GENE

hits = distanceToNearest(peak_gr, genes)

peak_to_gene = data.frame(
  peak=names(peak_gr)[queryHits(hits)],
  gene=genes$gene_name[subjectHits(hits)],
  gene_biotype = genes$gene_biotype[subjectHits(hits)],
  dist_to_gene=mcols(hits)$distance
)

head(peak_to_gene)
```

gene biotype: classification (so protein-coding, long noncoding, pseudogene)
```{r}
# merge gene annotations back onto DA table
# one table with stats (padj val, log2fc) and biology (gene name, biotype, distance)

sig_da_anno = sig_da %>% 
  left_join(peak_to_gene, by="peak") %>% 
  dplyr::select(peak, gene, gene_biotype, dist_to_gene,
                p_val_adj, avg_log2FC, pct.1, pct.2, p_val)

head(sig_da_anno)

# write to csv
write.csv(sig_da_anno, "DA_peaks_CHLA9_vs_CHLA10_annotated_nearest_gene.csv", row.names=FALSE)
```


*Label volcano plot with significant genes*

```{r}
library(ggplot2)
library(dplyr)
library(ggrepel)
# had to reload data from csv file because issues with RStudio, forgot to save .rds
lr_da_peaks2 = read.csv("DA_peaks_CHLA9_vs_CHLA10_scATAC.csv")
sig_da_anno  = read.csv("DA_peaks_CHLA9_vs_CHLA10_annotated_nearest_gene.csv")

# select labels
labels_df = sig_da_anno %>%
  filter(p_val_adj < 1e-30, abs(avg_log2FC) > 2) %>%
  mutate(
    score = -log10(p_val_adj),
    direction = ifelse(avg_log2FC > 0, "CHLA9", "CHLA10")
  ) %>%
  group_by(direction) %>%
  slice_max(order_by = score, n = 15) %>%   # can adjust
  ungroup() %>%
  distinct(peak, .keep_all = TRUE) %>%
  select(peak, gene)


plot_df = lr_da_peaks2 %>%
  mutate(Direction = case_when(
    p_val_adj <= 0.05 & avg_log2FC >= 0.25 ~ "More open in CHLA9",
    p_val_adj <= 0.05 & avg_log2FC <= -0.25 ~ "More open in CHLA10",
    TRUE ~ "Not sig"
  ))

volcano_genes = ggplot(plot_df, aes(x = avg_log2FC, y = -log10(p_val_adj), color = Direction)) +
  geom_point(alpha = 0.6, size = 1) +
  theme_classic(base_size = 18) +
  xlab("log2FC accessibility (CHLA9 − CHLA10)") +
  ylab("-log10(adj p-value)") +
  #ggtitle("Differential accessibility (ATAC): CHLA9 vs CHLA10") +
  geom_label_repel( # repel adds box around label vs geom_label_text
    data = plot_df %>% inner_join(labels_df, by = "peak"),
    aes(label = gene),
    size = 3.5,
    max.overlaps = Inf,
    box.padding = 0.6, # space around text
    point.padding = 0.8, # space around dot
    segment.size = 0.25,
    force=10,
    show.legend = FALSE
  ) +
  theme(
    legend.title=element_text(size=16),
        legend.text=element_text(size=14)) +
  guides(
    color=guide_legend(
      override.aes=list(size=3)
    )
  ) +
  scale_color_manual(
    name="Direction",
    values = c(
  "More open in CHLA10" = "#0072B2",
  "More open in CHLA9"  = "#C76E00",
  "Not sig"             = "grey80"
)
  )

volcano_genes
ggsave("DA_CHLA9_CHLA10_figure.svg", width=10, height=8, dpi=900)
```
*UNLABELLED VOLCANO PLOT*
Figure for poster presentation

```{r}

```

```{r}
# sanity check to make sure mapping worked
sum(is.na(sig_da_anno$gene))
summary(sig_da_anno$dist_to_gene)
```

```{r}
# add direction, log2fc more interpretable
sig_da_anno = sig_da_anno %>% 
  mutate(direction = ifelse(avg_log2FC > 0, "More open in CHLA9","More open in CHLA10"))

table(sig_da_anno$direction)
```

RECALL from earlier volcano plot:
  positive values (right), peaks mean more accessibility in CHLA9 (pre-treatment)
      condition one is "up" relative to condition two "down"
  negative values (left), peaks mean more accessibility in CHLA10 (post-treatment)
      condition one is "down" relative to condition two "up"
      
      
```{r}
# overlap with RNA DEGs
# this is from the CHLA9_CHLA10_scRNA.rmd
de_stress2 = read.csv("/Users/thejesnair/Desktop/KCGIP/EwS/DEG_CHLA9_vs_CHLA10_StressLike_all.csv")
de_mes2 = read.csv("/Users/thejesnair/Desktop/KCGIP/EwS/DEG_CHLA9_vs_CHLA10_MesenchymalLike_all.csv")

deg_genes = unique(c(de_stress2$gene, de_mes2$gene))

sig_da_anno = sig_da_anno %>% 
  mutate(is_RNA_DEG = gene %in% deg_genes) # is the nearest gene to this DA peak in my RNA DEG list? but RNA DEG is only coming from 2 programs that had the biggest proportion of cells in CHLA9,10 cell lines

table(sig_da_anno$is_RNA_DEG)
```

so the 563 false can be from other programs, since I limited it to 2 programs with biggest proportions

600 true: DA Peaks near genes that are in our RNA DEGs data
563 false: DA peaks near genes that are NOT IN RNA DEGs data
  however, "false" is not bad or meaningless
    simply can be that DEGs are in other transcriptional programs (neural, proliferative)
    genes have post-transcriptional regulation
    regulatory priming w/o RNA output
    distal enhancer effects (look at TF enrichment for this?)
```{r}
table(seu_chla$final_annotation[Cells(seu_chla) %in% Cells(subset(seu_chla, features = sig_da_anno$peak))])
table(sig_da_anno$is_RNA_DEG, sig_da_anno$gene_biotype)
```
Peaks that show both accessibility change and RNA expression change tend to regulate protein-coding genes

dataset dominated by stress + mesenchymal states
DA peaks occuring in those cells proportionally

protein coding dominates true overlaps, so peaks overlapping RNA DEGs are mostly near protein-coding genes not lncRNA artifacts

biologically interpretable regulation

```{r}
# ATAC->RNA bridge (still need to look at motifs though)
sig_da_anno %>%
  mutate(overlap = ifelse(is_RNA_DEG, "Peak near RNA DEG", "Peak not near RNA DEG")) %>%
  ggplot(aes(x = direction, fill = overlap)) +
  geom_bar(position = "fill") +
  ylab("Fraction of DA peaks") +
  xlab("") +
  theme_classic() +
  ggtitle("DA peaks near RNA DEGs (nearest-gene mapping)")
```
Among peaks that are differentially accessible, what fraction are near genes that are RNA DEGs (stress-like mesenchymal-like)?

Started with sig DA peaks (atac)
split by: more open in CHLA10, CHLA9
for each group asked is the nearest gene to these open regions in the RNA analysis results? (stress + mes programs)

so bars are showing:
of ALL DA peaks biased towards this program, what fraction are near genes that ALSO change at the RNA level?

substnatial fraction of regulatory changes are reflected at RNA level, BUT not all-- this is similar pre and post treatment

supports idea that regulatory remodeling is widespread, only a part of it translates to detectable transcriptional changes in the programs tested


stacked bar plot confirms ATAC + RNA signals are not random


*GeneScoreMatrix DA*

Goal: identify genes whose ATAC-derived gene activity differs b/w CHLA9 and CHLA10


```{r}
# sanity check
DefaultAssay(seu_chla) # currently ATAC

# switch assay to GeneScoreMatrix
DefaultAssay(seu_chla) = "GeneScoreMatrix"

# more sanity checks, want to make sure data inside is what we want
seu_chla[["GeneScoreMatrix"]]
Layers(seu_chla[["GeneScoreMatrix"]]) # should see counts and data
genes_test = head(rownames(seu_chla[["GeneScoreMatrix"]]), 4)
VlnPlot(seu_chla, features = genes_test, group.by = "cell_line", pt.size = 0)
```


```{r}
# differential gene activity

# running findmarkers again because DIFFERENT assay, so it feels repetitive but it's not
# genescorematrix represents continuous, gene level ATAC signal
gs_deg = FindMarkers(
  seu_chla,
  ident.1="CHLA9",
  ident.2="CHLA10",
  test.use="wilcox", # distributional comparison, is gene more active? 
  # aggregated accessibility around genes, not sparse peak presence, wilcoxon is appropriate here
  min.pct = 0.1,
  logfc.threshold = 0.0 # will filter down later
)
```

```{r}
# format table
gs_deg2 = gs_deg %>% 
  tibble::rownames_to_column("gene") %>% 
  dplyr::arrange(p_val_adj) %>% 
  dplyr::select(gene, p_val_adj, avg_log2FC, pct.1, pct.2, p_val)

head(gs_deg2)

write.csv(gs_deg2, "GeneScore_DEG_CHLA9_vs_CHLA10.csv", row.names=FALSE)

sum(gs_deg2$p_val_adj <=0.05)
head(gs_deg2, 10)
```


*Gene-centric convergence*

These are genes that are found in the RNA DEG analysis between CHLA9 and CHLA10
AND
Identified using genescore-differential between CHLA9 and CHLA10 (so gene level ATAC-derived activity)

So it's combining both transcriptional + regulatory signal

pct is just percentage of cells that show gene is accessible in

To summarize again:
1. RNA data (GeneExpressionMatrix)
- is this gene more expressed in CHLA9 or CHLA10?

2. ATAC peaks (ATAC)
- Is this specific DNA region more open in CHLA9 or CHLA10?

3. GeneScoreMatrix (ATAC -> gene)
- Is the chromatin around this gene more permissive in CHLA9 or CHLA10?
(regulatory potential, not expression)
- aggregates ATAC signal across a window around a gene
  - promoter, gene body
- collapses many peaks into one gene level value
- so ATAC-derived BUT gene-centric



The whole reason we're looking at genescorematrix is to look at the area AROUND the open region of chromatin and determine how permissive it is (is the area open and accessible for the cell to turn that gene on), then we can compare to the actual RNA DEG results and see if there's overlap and agreement 
```{r}
# gene-centric coverage (RNA DEGs + GeneScore DEGs)

# USE this from earlier
#deg_genes = unique(c(de_stress2$gene, de_mes2$gene))

# filtered significant GeneScore genes
gs_sig = gs_deg2 %>% 
  dplyr::filter(p_val_adj <= 0.05)

# overlap set between RNA DEG expression and significant GeneScore genes
convergent_genes = gs_sig %>% 
  dplyr::filter(gene %in% deg_genes)

nrow(convergent_genes)
head(convergent_genes, 20)

```
*Triple convergent table? (RNA + GeneScore + DA peak evidence)*

1. RNA DEG
 genes' expression changes
2. GeneScoreMatrix DEG
 area around chromatin is permissive (relative to other cell line, regulatory potential/accessibility)
 (>0, more accessible in CHLA9, <0, more permissive in CHLA10)
3. nearby DA peak
 local regulatory DNA that is open/closed
 
Identifies genes supported by RNA expression changes, gene-level chromatin accessibility, and ATAC DA peaks (directionality not assumed across all 3).

```{r}
colnames(de_stress2)
colnames(de_mes2)
```

Put direction (higher in which cell line?)
```{r}
convergent_genes2 = convergent_genes %>% 
  mutate(
    direction = ifelse(avg_log2FC>0, "Higher in CHLA9", "Higher in CHLA10"))

head(convergent_genes2)
convergent_genes2
write.csv(convergent_genes2, "ConvergentGENES_RNA_GeneScore_CHLA9_vs_CHLA10.csv", row.names=FALSE)
```

```{r}
# ALL THREE CONVERGING
triple_convergent = convergent_genes2 %>% 
  dplyr::filter(gene %in% sig_da_anno$gene) %>% 
  dplyr::rename(direction_genescore = direction)
triple_convergent
#nrow(convergent_genes); nrow(convergent_genes2)

write.csv(triple_convergent, "CHLA9_vs_CHLA10_RNA_ATAC_GeneScore_convergent_genes.csv", row.names=FALSE)
# all values are from genescore findmarkers() results
# 
```

Reduced from convergent_genes2 (4088) to triple_convergent (457), so these genes have
1. differential gene expression
  gene's expression changes b/w CHLA9 and CHLA10
2. permissive gene score (accessibility around chromatin regions)
  chromatin environment *AROUND* the gene is *MORE* accessible in one condition vs the other
3. open chromatin peaks in general
  specific chromatin region near gene is differentially accessible
 
*DONE WITH ATAC, GENESCOREMATRIX*


*Initial look into MotifMatrix*

May not have the full motif/peak backing needed for Signac's built in enrichment functions since ATAC assay is a normal Seurat assay and not ChromatinAssay

```{r}
# switch to MotifMatrix

# what is motif assay keyed by?
DefaultAssay(seu_chla) = "MotifMatrix"
head(rownames(seu_chla[["MotifMatrix"]]),5) # outputs TF_ names, transcription factors

# Are there any peak-level motif matrices stored?
seu_chla[["ATAC"]]
length(rownames(seu_chla[["ATAC"]]))

names(seu_chla[["ATAC"]]@misc)
```

```{r}
nrow(seu_chla[["ATAC"]]) # total num peaks 
nrow(lr_da_peaks) # identified after running FindMarkers()
nrow(lr_da_peaks2)

sum(lr_da_peaks$p_val_adj <= 0.05, na.rm = TRUE) # 1300 significant peaks
sum(lr_da_peaks$p_val_adj <= 0.05 & abs(lr_da_peaks$avg_log2FC) >= 0.25, na.rm = TRUE) # 1163 after filtering

```

```{r}
# clearing
#rm(list=ls(all.names=TRUE))
#gc()
#ls()
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

