---
title: "scRNA 9cl GO/GSEA"
author: "Thejes Nair"
date: "2025-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(tidyverse)
library(clusterProfiler)
library(msigdbr)
library(enrichplot)
library(org.Hs.eg.db)
library(dplyr)
library(purrr)
library(ggplot2)
#library(presto) # issues installing locally
```

GSEA vs GO

GO (gene ontology enrichment)
only use significant DEGs (so meet a certain threshold)
"what pathways are overrepresented amongst the genes that are *significantly* up/down in this cluster?*

- intuitive
- good for top-upregulated/downregulated genes
- useful for cluster identity

* requires arbitrary thresholds
misses weak/biological coordinated pathways
up and down need to be run separately

GO great for annotation

GSEA (gene set enrichment analysis)
uses ALL genes ranked by log2FC-so no thresholds
even nonsig genes will contribute

"do coordinated changes across many genes in a pathway collectively go up or down?"

- no threshold bias
- catches subtle but real processes
- works with hallmark pathways (like EMT, G2M checkpoint)
- more biologically reliable

* can be slower
* harder to quickly summarize

```{r}
# load in objects
seurat_rnaonly = readRDS("/Users/thejesnair/Desktop/KCGIP/EwS/seurat_rna_only_forGSEA_GO.rds")
head(seurat_rnaonly)

# check default assay
DefaultAssay(seurat_rnaonly) # GeneExpressionMatrix

# for GSEA/GO want to use the clusters as the identities
Idents(seurat_rnaonly) = seurat_rnaonly$seurat_clusters
levels(Idents(seurat_rnaonly))
```

```{r}
# GOING TO DO GSEA LATER HAVE TO RERUN TO GET ALL RANKED GENES, NOT JUST TOP ONES
# get DEGS for all 9 clusters
# want all the genes with logFC (up- and down-)
deg_all = FindAllMarkers(
  seurat_rnaonly,
  assay = "GeneExpressionMatrix",
  only.pos = FALSE,
  min.pct = 0,
  logfc.threshold = 0
)

# add gene symbols as an explicit col
rownames(deg_all)
```

```{r}
# GO PATHWAY ANALYSIS
cluster_markers = readRDS("/Users/thejesnair/Desktop/KCGIP/EwS/cluster_markers.rds")
head(cluster_markers)
str(cluster_markers)

# keep only the significant DEGs
deg_sig = cluster_markers %>% 
  filter(p_val_adj <= 0.05)

# keep UP- and DOWN-regulated genes for GO (log2fc > 0)
up_deg = deg_sig %>% 
  filter(avg_log2FC > 0)

down_deg = deg_sig %>% 
  filter(avg_log2FC < 0)

table(up_deg$cluster)
table(down_deg$cluster)
```

```{r}
# CONVERT GENE SYMBOLS -> ENTREZ IDS (clusterprofiler)
gsym_entrez_up = bitr(
  up_deg$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

gsym_entrez_down = bitr(
  down_deg$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# reattach cluster col
deg_up_entrez = up_deg %>% 
  inner_join(gsym_entrez_up, by = c("gene" = "SYMBOL"))

deg_down_entrez = down_deg %>% 
  inner_join(gsym_entrez_down, by = c("gene" = "SYMBOL"))

# check
table(deg_up_entrez$cluster) # reduced slightly so must be genes with no symbol or no proper name?
table(deg_down_entrez$cluster)

```

```{r}
# building a list of gene sets for each cluster
up_gene_set = deg_up_entrez %>% 
  group_by(cluster) %>% 
  summarize(ENTREZID= list(unique(ENTREZID)), .groups="drop")

down_gene_set = deg_down_entrez %>% 
  group_by(cluster) %>% 
  summarize(ENTREZID= list(unique(ENTREZID)), .groups="drop")

up_gene_set
down_gene_set
```
```{r}
# run GO on each cluster (UP genes)
up_list = setNames(up_gene_set$ENTREZID,
                   up_gene_set$cluster)

# run GO BP enrichment for each cluster uP genes
go_up_list = lapply(
  names(up_list),
  function(cl){
    enrichGO(
      gene = up_list[[cl]],
      OrgDb = org.Hs.eg.db,
      keyType = "ENTREZID",
      ont = "BP", #biological process
      pAdjustMethod = "BH",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2,
      readable = TRUE
    )
  }
)

names(go_up_list) = names(up_list)

head(go_up_list[["0"]])
```

```{r}
# DOWN regulated genes, GO
down_list = setNames(down_gene_set$ENTREZID,
                   down_gene_set$cluster)

# run GO BP enrichment for each cluster DOWN genes
go_down_list = lapply(
  names(down_list),
  function(cl){
    enrichGO(
      gene = down_list[[cl]],
      OrgDb = org.Hs.eg.db,
      keyType = "ENTREZID",
      ont = "BP", #biological process
      pAdjustMethod = "BH",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2,
      readable = TRUE
    )
  }
)

names(go_down_list) = names(down_list)

head(go_down_list[["0"]])

```

```{r, fig.height=9,, fig.width=7}
# dotplot visualizations

# upregulated GO BP
dir.create("GO_plots", showWarnings = FALSE)

for (cl in names(go_up_list)) {
  go = go_up_list[[cl]]
  if (is.null(go) || nrow(as.data.frame(go)) == 0) next  # skip empty
  
  p = dotplot(go, showCategory = 15) +
    ggtitle(paste("Cluster", cl, "- GO BP (UP genes)")) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.y = element_text(size = 10)
    )
  
  print(p)  # visualize
  
  ggsave(
    filename = file.path("GO_plots", paste0("GO_UP_cluster", cl, ".png")),
    plot = p,
    width = 7, height = 5, dpi = 300, bg = "white"
  )
}

```



```{r, fig.height=9, fig.width=7}
# downregulated GO BP
for (cl in names(go_down_list)) {
  go = go_down_list[[cl]]
  if (is.null(go) || nrow(as.data.frame(go)) == 0) next  # skip empty
  
  p = dotplot(go, showCategory = 15) +
    ggtitle(paste("Cluster", cl, "- GO BP (DOWN genes)")) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.y = element_text(size = 10)
    )
  
  print(p)
  
  ggsave(
    filename = file.path("GO_plots", paste0("GO_DOWN_cluster", cl, ".png")),
    plot = p,
    width = 7, height = 5, dpi = 300, bg = "white",
  )
}

```

```{r}
# want a table of cell line distribution across clusters
seurat_integration = readRDS("/Users/thejesnair/Desktop/KCGIP/EwS/seurat_integration_finalAnnotations.rds")


# bar plot?
tab = table(seurat_integration$cell_line, seurat_integration$final_annotation)

# convert to proportions within each cluster
prop_tab = prop.table(tab, margin = 2)  # normalize by columns (clusters)

library(pheatmap)

cl_heatmap = pheatmap(
  prop_tab,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("white", "darkblue"))(80),
  fontsize_row = 12,
  fontsize_col = 12,
  main = "Proportion of Each Cell Line in Each Cluster"
)

ggsave(cl_heatmap, filename = "cl_heatmap.png")

```

```{r}
FeaturePlot(seurat_integration, features = c("HIVEP3", "SUCLA2"))
VlnPlot(seurat_integration, features = c("HIVEP3", "SUCLA2"), group.by = "cell_line")
DotPlot(seurat_integration, features = c("HIVEP3", "SUCLA2"), group.by="cell_line")
```
*GSEA*
* Ran presto on Talapas apptainer because having issues installing locally *

*ON TALAPAS*
```{r setup, include=FALSE}
library(Seurat)
library(presto)
library(dplyr)
library(purrr)
library(msigdbr)

sessionInfo()
```

```{r}
# load integrated obj
seurat_integration_finalAnno = readRDS("/projects/bgmp/shared/groups/2025/sarcoma/tnair/Seurat/seurat_integration_finalAnnotations.rds")

# Find all markers (all genes now) using PRESTO (so much faster)
allmarkers_presto = FindAllMarkers(
  seurat_integration_finalAnno,
  assay = "GeneExpressionMatrix",
  only.pos = FALSE,
  logfc.threshold = 0, # keep ALL genes
  min.pct = 0)  
```

```{r}
# save obj
saveRDS(allmarkers_presto, "allmarkers_presto.rds")

#  slim down to what is needed for GSEA
deg_all = allmarkers_presto %>% 
  dplyr::select(gene, cluster, avg_log2FC, p_val_adj)

# save obj
saveRDS(deg_all, file = "deg_all_GSEA.rds")
```

*LOCALLY* 

```{r}
# load deg_all obj from presto output
deg_all = readRDS("/Users/thejesnair/Desktop/KCGIP/EwS/deg_all_GSEA.rds")
head(deg_all)
```

```{r}
# build ranked gene lists per cluster
make_geneList = function(df) {
  df %>% 
    mutate(stat = avg_log2FC) %>% # ranking metric
    arrange(desc(stat)) %>% ## highest positive first
    distinct(gene, .keep_all = TRUE) %>% # one row per gene
    pull(stat, name = gene) # named numeric vector
}

geneLists = split(deg_all, deg_all$cluster) %>% 
  lapply(make_geneList)

head(geneLists[["0"]])
```

```{r}
# MSigDb gene sets (Hallmark + GO BP)

# Hallmark
msig_h = msigdbr(
  species = "Homo sapiens",
  collection = "H"
) %>% 
  dplyr::select(gs_name, gene_symbol)

# GO BP
msig_c5_bp = msigdbr(
  species = "Homo sapiens",
  collection = "C5",
  subcollection = "BP"
) %>% 
  dplyr::select(gs_name, gene_symbol)

# combine the table, term2gene
term2gene = dplyr::bind_rows(msig_h, msig_c5_bp)
```

```{r}
# GSEA per cluster
gsea_res = lapply(names(geneLists), function(cl) {
  gl = geneLists[[cl]]
  message("Running GSEA for cluster ", cl, " (n genes = ", length(gl), ")")
  
  GSEA(geneList = gl,
       TERM2GENE = term2gene,
       pAdjustMethod = "BH",
       minGSSize = 10,
       maxGSSize = 500,
       pvalueCutoff = 0.05,
       verbose = FALSE)
})

names(gsea_res) = names(geneLists)
```

```{r, fig.height = 10, fig.width = 10}
# sanity check

g2_df = as.data.frame(g2)

dim(g2_df)
head(g2_df)

plot_gsea = function(gsea_res,
                     outdir = "GSEA_plots",
                     prefix = "GSEA",
                     n_category = 20,
                     width = 10, height = 10) {
  # make output dir if not exists
  dir.create(outdir, showWarnings = FALSE)
  
  for (cl in names(gsea_res)) {
    gsea_obj = gsea_res[[cl]]  # get result for this cluster
    
    
    p = dotplot(gsea_obj, showCategory = n_category) +   # dotplot on gsea obj
      ggtitle(paste("Cluster", cl, "-", prefix)) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.y = element_text(size = 10)
      )
    
    print(p)
    
    outfile = file.path(outdir, paste0(prefix, "_cluster", cl, ".png"))
    ggsave(outfile, plot = p, width = width, height = height,
           dpi = 300, bg = "white")
    
    message("Saved: ", outfile)
  }
}

plot_gsea(
  gsea_res = gsea_res,
  outdir   = "GSEA_plots",
  prefix   = "Hallmark_GO",
  n_category = 20
)
```

```{r, fig.width=13, fig.height = 8}

umap_ppt =  DimPlot(
  seurat_integration,
  reduction = "umap",
  group.by = "final_annotation",
  label = FALSE
) + labs(x="UMAP1", y="UMAP2") + ggtitle("") + theme_bw() + theme(
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.line = element_line(color="black", linewidth=0.3),
    legend.text = element_text(size = 18),
    legend.justification = "center",
    legend.key.spacing.y = unit(0.3, "cm")
)

umap_ppt

ggsave(filename = "umap_ppt.png", plot = umap_ppt, 
       dpi = 900, bg = "white", width = 13,
       height = 8, units = "in")
```

```{r}
# GSEA cluster 1
g1 = as.data.frame(gsea_res[["1"]])

g1_adhesion = g1 %>% 
  dplyr::filter(grepl("ADHESION|EXTRACELLULAR MATRIX|FOCAL ADHESION|MIGRATION",
                      Description, ignore.case = TRUE))

g1_adhesion[, c("ID", "Description", "NES", "p.adjust")] %>% head(20)

g2 = as.data.frame(gsea_res[["2"]])
g2_adhesion = g2 %>% 
  dplyr::filter(grepl("ADHESION|EXTRACELLULAR MATRIX|FOCAL ADHESION|MIGRATION",
                      Description, ignore.case = TRUE))

g1_adhesion
```







