---
title: "scRNA 9cl program annotation"
author: "Thejes Nair"
date: "2025-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(Seurat)
library(dplyr)
library(patchwork)
library(tidyverse)
library(readr)
library(pheatmap)
library(ggplot2)
library(ggrepel)
library(future)
library(msigdbr)
library(scCustomize)
```

*Clean up genes by cell line data, rearrange*

```{r}
# load tidyverse to go through data
raw.lines = read_lines("/projects/bgmp/shared/groups/2025/sarcoma/shared/filtered_struc_variants/celllinesgenes.tsv")
head(raw.lines)

# initialize vectors to store results
cell.lines.vec = c()
genes.vec = c()

#EwS 9 cell lines
ews.cls = c("A673", "A4573", "CHLA9", "CHLA10", "PDX305", 
            "RDES", "SKNMC", "TC32", "TC71")
# variables
current.cellline = NULL #placeholder for current cell line

# loop through each line
for (line in raw.lines) {
 clean.line = str_trim(line) #remove trailing space
 if (clean.line %in% ews.cls) {
   current.cellline = clean.line #Cell line
 } else {
   #Gene name
   cell.lines.vec = c(cell.lines.vec, current.cellline)
   genes.vec = c(genes.vec, clean.line)
 }
}
```

```{r}
# combine cls and genes into df
df = tibble(cell_line = cell.lines.vec, gene = genes.vec)

# write to tsv
write_tsv(df, "genes_by_cellline_clean.tsv")

print(head(df))
print(tail(df))
```

```{r}
# turn df into vectored list for each cell

genes_celline = df %>% 
  distinct(cell_line, gene) %>% 
  group_by(cell_line) %>% 
  reframe(genes=list(unique(gene))) %>% deframe() #return tibble to list
```

```{r}
# check df
class(genes_celline)
names(genes_celline)
genes_celline$A673
```

*SEURAT OBJECT*

```{r}
# load Seurat obj, cross fingers and toes
rm(list = ls()); gc() #clear memory bc otherwise get an error if i try to reload data
seurat_ob = readRDS("/projects/bgmp/shared/groups/2025/sarcoma/shared/Multiome/9cl_seurat.rds")
exists("seurat_ob")
```
Let's do some EDA and take a look more into Seurat object/metadata
```{r}
seurat_ob
```
Default assay is MotifMatrix --> need to change to GeneExpressionMatrix (scRNA-seq)
Outlined more in lab notebooks

```{r}
DefaultAssay(seurat_ob) = "GeneExpressionMatrix" # set default/active assay to scRNA-seq data
slotNames(seurat_ob[["GeneExpressionMatrix"]]) # what's stored inside the assay
VariableFeatures(seurat_ob)[1:5] 
colnames(seurat_ob@meta.data) # check for clustering
```
counts: raw counts
data: log normalized values, from NormalizeData()
scale.data: scaled expression (used for PCA), from ScaleData()
assay.orig: made when seurat objs are saved/merged
var.features: highly variable genes identified from FindVariableFeatures() used for downstream analysis like PCA

**VariableFeatures**: most variable genes across cells in dataset, expression fluctuates after normalization
But this is all before we've integrated

Code that did these steps from github script:
https://github.com/furlan-lab/EwS_multiome/blob/main/code/make_object_9_ewing_cell_lines.Rmd

seu<-NormalizeData(t) %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA()

```{r}
# integrated and clustered?
"integrated" %in% Assays(seurat_ob)
names(seurat_ob@commands) # look at command history
# no "FindIntegrationAnachors" or "IntegrateLayers" "IntegrateData" 
head(seurat_ob@meta.data) # META DATA 
Reductions(seurat_ob)# what dimensionality reduction techniques were done (PCA, UMAP)

seurat_ob[["pca"]]@assay.used # verifying that PCA and UMAP dimensionality reduction techniques done on UNINTEGRATED data
seurat_ob[["mo_UMAP"]]@assay.used
```
mo_UMAP and pca reductions
pca, mo_UMAP: pc based on RNA only (GeneExpressionMatrix)
thought mo_UMAP was with combined sc data, it isn't
also thought this meant data was already integrated, but also no

*INTEGRATE*

```{r}
unique(seurat_ob@meta.data$cell_line) # making sure cell lines are in this col

# SPLIT object by cell lines
objs = SplitObject(seurat_ob, split.by = "cell_line")

# data is already normalized, there are already variable features per object (this will change though after integration)
# picking integration features
features = SelectIntegrationFeatures(object.list = objs, nfeatures = 2000 ) #nfeatures = 2000 from vignettes

# check that obj split successfully 
class(objs) # list
length(objs) # 9 cell lines
```

```{r}
# find anchors using CCA
anchors = FindIntegrationAnchors(object.list = objs,
                                 normalization.method = "LogNormalize",
                                 anchor.features = features,
                                 reduction = "cca")
```

```{r}
class(anchors)
anchors
```

```{r}
# integrate, make sure to set normalization method to what was used previously and use features created
seurat_integration = IntegrateData(anchorset = anchors,
                                   normalization.method = "LogNormalize",
                                   features.to.integrate = features,
                                   new.assay.name = "integratedCCA")
```

```{r}
# confirm integration
class(seurat_integration)

# does assay exist?
Assays(seurat_integration)
Key(seurat_integration[["integratedCCA"]]) # will end with _

seurat_integration
```
*PCA, UMAP, CLUSTERING, VISUALIZATIONS*

```{r}
DefaultAssay(seurat_integration) = "integratedCCA"

# standard Seurat workflow for clustering and visualization
seurat_integration = ScaleData(seurat_integration, verbose = FALSE)
seurat_integration = RunPCA(seurat_integration, npcs=30, verbose = FALSE)

# before creating UMAP visuals, check the PC dimensions, set it to 30 earlier but will probably be much less
ElbowPlot(seurat_integration, ndims=30) # cut off around 18

set.seed(31) # consistent UMAP orientation in re-run
seurat_integration = RunUMAP(seurat_integration, dims=1:18, verbose= FALSE) #use PCA dimensions for UMAP
seurat_integration = FindNeighbors(seurat_integration, dims=1:18)
seurat_integration = FindClusters(seurat_integration, resolution = 0.5)
```
Quick checks

```{r}
# sanity check, built on integrated assay?
seurat_integration[["pca"]]@assay.used
"integratedCCA" %in% Assays(seurat_integration)

# clusters present?
colnames(seurat_integration@meta.data) #seurat_clusters !!!!, 'Clusters' is from original unintegrated data
head(Idents(seurat_integration))
levels(Idents(seurat_integration)) # 8 clusters

```

Went from 25 clusters (unintegrated) to 9 after integration (clusters 0-8) so redundant, dataset-specific clusters were collapsed. Integration alignedshared cellular states/transcriptional programs amongst cell lines, whereas unintegrated was just mapping everything--batch effects, noise, dataset specific differences, which was increasing the number of clusters. Integration leaves biological heterogeneity behind, giving a better reflection of the transcriptional diversity in the 9 cell lines.

*SAVE INTEGRATED RDS OBJECT*

```{r}
saveRDS(seurat_integration, file = "integrated_seurat.rds")
```

```{r}
# load file for later us
seurat_integration = readRDS("/projects/bgmp/shared/groups/2025/sarcoma/tnair/Seurat/integrated_seurat.rds")
```

*checking everything is good for Vini*

```{r}
# is an object
class(seurat_integration)

seurat_integration[["GeneExpressionMatrix"]]

# metadata, cell lines + clusters (seurat_clusters is integrated clusters)
colnames(seurat_integration@meta.data)
head(seurat_integration@meta.data)

# what assays
Assays(seurat_integration) # integratedCCA from me

# which reductions
Reductions(seurat_integration)

# seurat clusters contain info
table(seurat_integration$seurat_clusters)  # yes they do

seurat_integration[["GeneExpressionMatrix"]]@data[1:5, 1:5]
seurat_integration[["GeneExpressionMatrix"]]@counts[1:5, 1:5]
seurat_integration[["GeneExpressionMatrix"]]@data["POLQ", ] %>% head()
VlnPlot(seurat_integration, features = "POLQ", group.by = "cell_line")
dim(seurat_integration[["GeneExpressionMatrix"]]@data) # 36,438 genes × 19,737 cells
```
```{r}
DimPlot(seurat_integration,
        group.by = orig.ident)

colnames(seurat_integration@meta.data)
seurat_ob@Ass
```



```{r}
head(seurat_integration)
seurat_integration

D
```



*UMAP VISUALIZATION*

```{r}
# UMAP colored by clusters
p1 = DimPlot(seurat_integration, reduction = "umap", group.by="seurat_clusters", label=TRUE, repel=TRUE) + ggtitle("EwS Clusters")
p2 = DimPlot(seurat_integration, reduction = "umap", group.by="cell_line", label=TRUE, repel=TRUE) + ggtitle("EwS Cell lines")

combined_UMAPS = p1 + p2

combined_UMAPS

ggsave("combined_UMAPS.jpg", plot=combined_UMAPS, device = "jpeg", width = 13, height = 7, dpi = 300, bg = "white")
```

Several clusters (3,6, and 7) are predominantly composed of one cell line, SKNMC (retroorbital metastasis), whereas other clusters contained cells from multiple lines. Dominant clusters likely represent cell line–specific transcriptional programs that persist even after integration, while the mixed clusters correspond to shared EwS cellular states.


Some cell lines are clustering together (indicated by line connecting labels) so want to see if I can quantify that

```{r}
# clusters (columns) by cell lines (rows)
tab = table(seurat_integration$cell_line, Idents(seurat_integration))
tab

# convert to proportions within each cluster
prop_tab = prop.table(tab, margin = 2)
round(prop_tab, 2)

# proportion of cell line's cells in that specific cluster
library(pheatmap)
line_vs_cluster = pheatmap(prop_tab, cluster_rows = FALSE, cluster_cols = FALSE,
         color = colorRampPalette(c("white", "darkblue"))(50),
         #main= "Proportion of Cell Lines in Clusters",
         fontsize_row = 12,
         fontsize_col = 12,
         fontsize = 14)

line_vs_cluster

ggsave("cell_line_vs_cluster.png", plot=line_vs_cluster, width = 13, height = 7, dpi = 300)
ggsave("cell_line_vs_cluster2.jpg", plot=line_vs_cluster, device = "jpeg", width = 13, height = 7, dpi = 300, bg = "white")

```

*FIND MARKERS (TOP UPREGULATED/DOWNREGULATED GENES)*

We want the integrated clusters as the classifying variable for top marker identification
Take cells grouped by integrated clusters (shared biological states) and compare gene expression values to see what's differentially expressed -> finding genes that DISTINGUISH integrated clusters -> later on annotate clusters


```{r}
seurat_integration # default assay is integratedCCA, need to switch to geneexpressionmatrix

# utilizing seurat integration obj that contains integratedCCA assay, but switching to RNA data for marker identification
# maintain cluster ids and umap coordinates

DefaultAssay(seurat_integration) = "GeneExpressionMatrix"
#seurat_integration # switched default

# making sure classifying variable is set to integrated seurat_clusters, not Clusters from original obj
#head(Idents(seurat_integration)) # 0-8 clusters
#all(Idents(seurat_integration) == seurat_integration$seurat_clusters) # TRUE
#all(Idents(seurat_integration) == seurat_integration$Clusters) # FALSE
```

*top markers*

```{r}
plan("multicore", workers = 12) # set up for parallelization
markers = FindAllMarkers(seurat_integration,
                         assay="GeneExpressionMatrix",
                         logfc.threshold = 0.25, # default in older version (current is 0.1)
                         min.pct = 0.1, # default in older version/HBC training (current is 0.01) gene expressed in >10% of cells
                         only.pos = FALSE) # both up/down regulated

plan("sequential") # return to default

# can later subset markers to find pct w/ greater values pct.1, pct.2
#strict_markers = markers %>% filter(pct.1 >= 0.25, (pct.1 - pct.2) >= 0.1)

```

Note: find all markers takes a while, took ~1hr

**cluster:** cluster where gene is enriched
**avg_log2FC:** log2 FC of avg expression b/w cluster and all others (higher=more strongly upregulated, lower=more strongly downregulated)
**pct.1:** fraction of cells in this cluster that express gene
  so 0.75, 75% of cells in this cluster express this gene
**pct.2:** fraction of cells in OTHER clusters that express this gene
  so 0.10, 10% cells outside o cluster express
**p_val:** raw p-val (wilcoxon)
**p_val_adj:** adj p val correcting for multiple comparisons
  control for false positives
  
Upregulated: fusion-driven activation signatures(potential targets of EWS:FLI1?)

Downregulated: repressed developmental/differentation genes (targets repressed by EWS:FLI1?)
  mesenchymal programs silenced?

EWSLLFLI1 is unique bc it acts as activator and supressor 
Variant analysis later on will determine where there's overexpression (amplification), loss of expression (deleiton/truncation)
Top/down markers will be useful for variant integration and pathway analysis later on as well (pathway/GO enrichment)

```{r}
# look at data
head(markers)

# rearrange
markers = markers %>% 
  select(gene, cluster, avg_log2FC, p_val_adj, pct.1, pct.2, p_val)

head(markers)
colnames(markers)
#write.csv(head(markers,5), "test.csv", row.names = FALSE)

# sig vs nonsig?
markers %>%
  summarize(
    total_genes = n(),
    significant = sum(p_val_adj <= 0.05),
    nonsignificant = sum(p_val_adj > 0.05)
  )

# filter sig genes, subset top up/down regulated

sig_markers = markers %>% filter(p_val_adj <= 0.05)

up_reg_markers = sig_markers %>% 
  filter(avg_log2FC > 0) %>% # log2FC GREATER than 0
  group_by(cluster) %>% 
  slice_max(order_by= avg_log2FC, n = 100) # selects top 100 genes PER cluster

down_reg_markers = sig_markers %>% 
  filter(avg_log2FC < 0) %>% # log2FC LESS than 0
  group_by(cluster) %>% 
  slice_min(order_by = avg_log2FC, n = 100)
```
```{r}
# peek into top 100 genes up/down
head(up_reg_markers)
head(down_reg_markers)

# 100 per cluster? so 900 total
table(up_reg_markers$cluster)
table(down_reg_markers$cluster)
```

*DELIVERABLES FOR ENRICHMENT/PATHWAY ANALYSES*

```{r}
# save integrated marker object
head(markers)
saveRDS(markers, "cluster_markers.rds")

# save csv files for Josh
write.csv(up_reg_markers, "top100_upregulated_markers.csv", row.names = FALSE)
write.csv(down_reg_markers, "top100_downregulated_markers.csv", row.names = FALSE)
```

```{r}
# load integrated markers
markers = readRDS("/projects/bgmp/shared/groups/2025/sarcoma/tnair/Seurat/cluster_markers.rds")
class(markers)
head(markers)
```


```{r}
seurat_ob # original obj
```

```{r}
seurat_integration # integrated obj
Layers(seurat_integration[["GeneExpressionMatrix"]])   # expect counts, data (and scale.data if scaled)
Layers(seurat_integration[["integratedCCA"]])          # data (and scale.data if scaled)
dim(seurat_integration)
head(seurat_integration)
```
Determine which cell line is enriched in which cluster
```{r}
# quick breakdown
table(seurat_integration$seurat_clusters, seurat_integration$cell_line)
```
Which clusters correspond to which cell lines
```{r}
# more comprehensive
seurat_integration@meta.data %>%
  group_by(seurat_clusters, cell_line) %>%
  summarize(n = n()) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  arrange(seurat_clusters, desc(percent))
```

```{r}
# creates small UMAP per cell line, which clusters are present/enriched in each cell line?
clusters_cellline = DimPlot(seurat_integration, group.by = "seurat_clusters", split.by = "cell_line")
clusters_cellline
```
*SELECTING GENES FOR ANNOTATING CANONICAL PATHWAY*
```{r}
# VlnPlot(seurat_integration, features = "VEGFA")
# FeaturePlot(seurat_integration, features = "VEGFA")
```

```{r}
# load in HOXD13 KD up/down regulated genes from Apfelbaum 2022 paper (supplmenetal table s3, NIHMS1814671-supplement-4)

HOXD13_table = read.csv("/projects/bgmp/shared/groups/2025/sarcoma/tnair/HOXD13_KD_regulated_genes.csv",header=TRUE)
head(HOXD13_table)

# neural-like
neural_like_genes = HOXD13_table$Gene.name[HOXD13_table$State == "Neural-like"]

# mesenchymal-like
mes_like_genes = HOXD13_table$Gene.name[HOXD13_table$State == "Mesenchymal-like"]

```

```{r}
cl0_up = up_reg_markers %>% filter(cluster == 0)
cl0_up_genes = cl0_up$gene
length(cl0_up_genes)

# compare c0 upreg genes to HOXD13 programs
neural_overlap_c0 = intersect(cl0_up_genes, neural_like_genes)
mes_overlap_c0 = intersect(cl0_up_genes, mes_like_genes)
length(neural_overlap_c0) # 0
length(mes_overlap_c0) # 1

# no meaningful neural vs mesenchymal programs

# extract c0 top DEGs
```

```{r}
# go through all clusters, upreg genes, try and assign neural-like, mesenchymal-like programs
# ambiguous/clusters w/o assignment will tackle later on

# upreg genes by cluster
cluster_up_list = split(up_reg_markers$gene, up_reg_markers$cluster)
head(cluster_up_list[["0"]]) # genes for cluster 0
length(cluster_up_list[["0"]]) # 100 genes
```

```{r}
# loop to see overlap counts per cluster
program_results = up_reg_markers %>% 
  group_by(cluster) %>% 
  summarize(
    neural_hits = sum(gene %in% neural_like_genes),
    mes_hits = sum(gene %in% mes_like_genes),
    .groups = "drop"
  )

program_results

# annotation table
# signal strengths are RELATIVE to other clusters

annotations = program_results %>% 
  mutate(
    "HOXD13 signal" = case_when(
      neural_hits >=3 ~ "strongest neural signal",
      neural_hits == 1 ~ "weaker neural",
      mes_hits >= 2 ~ "moderate mesenchymal",
      mes_hits ==1 ~ "weaker mesenchymal",
      TRUE ~ "none"
    ),
    "Possible annotations" = case_when(
      neural_hits >=3 ~ "Neural-like",
      neural_hits == 1 ~ "Possibly neural-like",
      mes_hits >=2 ~ "Mesenchymal-like",
      mes_hits == 1 ~ "Possibly mesenchymal-like",
      TRUE~"Ambiguous/other programs"
      
    )
  )
annotations
```

*POSSIBLE ANNOTATIONS*

*NEURAL*
Cluster 3: Neural-like
5,8: Weak neural

*MESENCHYMAL*
1: Moderate mes
0,6: Weak mes

*NEITHER*
2,4,7: ambiguous


```{r}
# making new obj so dont overwrite prev one
obj = seurat_integration

# make sure chars
#is.character(obj$seurat_clusters)
obj$seurat_clusters = as.character(obj$seurat_clusters)
annotations$cluster = as.character(annotations$cluster)

# annotation cols into vectors
signal_vec = annotations$'HOXD13 signal'
names(signal_vec) = annotations$cluster

state_vec = annotations$'Possible annotations'
names(state_vec) = annotations$cluster

# map onto cluster
mapped_signal = signal_vec[obj@meta.data$seurat_clusters]
mapped_state = state_vec[obj@meta.data$seurat_clusters]

obj@meta.data$HOXD13_signal = mapped_signal
obj@meta.data$HOXD13_state = mapped_state

table(obj@meta.data$HOXD13_state)
head(obj@meta.data[, c("seurat_clusters", "HOXD13_signal", "HOXD13_state")]) # now annotations will be alognside clusters

seurat_integration = obj

```

```{r}
# umap with new annotations
DimPlot(
  seurat_integration,
  group.by = "HOXD13_state",
  label = TRUE,
  repel = TRUE,
  pt.size = 0.3
) + ggtitle("HOXD13-based mesenchymal vs neural")

DimPlot(
  seurat_integration,
  group.by = "HOXD13_signal",
  label = TRUE,
  repel = TRUE,
  pt.size = 0.3
)

p1
```

```{r}

# HOXD13 Programs

# UMAP coordinates + metadata into one data.frame
emb = Embeddings(seurat_integration, "umap")
emb = as.data.frame(emb)
emb$cluster = seurat_integration$seurat_clusters
emb$HOXD13_state = seurat_integration$HOXD13_state
emb$cell_line = seurat_integration$cell_line
emb$Phase = seurat_integration$Phase

# center of each data to place labels
centers = aggregate(
  cbind(umap_1, umap_2) ~ cluster,
  data = emb,
  FUN = mean
)

# plot points colored by HOXD13_state, labels are cluster IDs
hoxd13state_plot = ggplot(emb, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(color = HOXD13_state), size = 0.3, alpha = 0.8) +
  geom_text(
    data = centers,
    aes(label = cluster),
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    title = "Clusters by HOXD13 programs",
    color = "HOXD13_state"
  ) +
  guides(
    color = guide_legend(
      override.aes = list(size = 2) # increase dot size
    )
  )


(hoxd13state_plot | p2) / clusters_cellline

hoxd13state_plot
```
*CELL CYCLE SCORING*

Seurat has CellCycleScoring that assigns cells to diff cell cycle phases G1, S, or G2/M phases


```{r}

# want to do cell cycle scoring on RAW data!
# before default was integratedCCA
DefaultAssay(seurat_integration) = "GeneExpressionMatrix"

# seurat 2019 updated gene sets
s_genes = cc.genes.updated.2019$s.genes
g2m_genes = cc.genes.updated.2019$g2m.genes

# run cell cycle scoring on INTEGRATED OBJ
seurat_integration = CellCycleScoring(
  object = seurat_integration,
  s.features = s_genes,
  g2m.features = g2m_genes,
  set.ident = FALSE # so Idents isn't overwritten
)

# how many cells each phase?
table(seurat_integration$Phase)

```

```{r}
# switch back to integrated assay for visualizations
DefaultAssay(seurat_integration) = "integratedCCA"

# visualize, UMAP colored by cell-cycle phase
DimPlot(
  seurat_integration,
  reduction = "umap",
  group.by = "Phase",
  pt.size = 0.3
) + ggtitle("Cell cycle phase")

# cross reference cluster and phase
seurat_integration$seurat_clusters = as.character(seurat_integration$seurat_clusters)

# counts
table(seurat_integration$seurat_clusters, seurat_integration$Phase)

# proportions per cluster
prop.table(
  table(seurat_integration$seurat_clusters, seurat_integration$Phase),
  margin = 1  # row-wise
)

```

```{r}
# cell cycle phase and cluster
ggplot(emb, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(color = Phase), size = 0.25, alpha = 0.8) +
  geom_text(
    data = centers,
    aes(label = cluster),
    color = "black",
    fontface = "bold",
    size = 4
  ) +
  theme_classic() +
  labs(
    title = "Cell cycle phase with cluster labels",
    color = "Phase"
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)))

```
*CELL CYCLE STAGES*
0: G1, WEAK MES
1: S, MODERATE MES
2: G2/M, AMBIGUOUS
3: mixed, NEURAL-LIKE
4: mixed, AMBIGUOUS
5: G1, WEAK NEURAL
6: S, WEAK MES
7: mixed, AMBIGUOUS
8: G2/M, WEAK NEURAL

Proliferative: 1,2,6,8
Low proliferation: 0,5
Mixed: 3,4,7

--- 
Confident programs:
3: mixed, neural-like
1: S, moderate mes

Program exists, but weaker (remember this is all relative to other clusters):
* weaker but not necessarily totally ambiguous(?)
* provisionally labelled
0: G1, weak mes
5: G1, weak neural
6: S, weak mes
8: G2/M, weak neural

Ambiguous:
2: G2/M, ambiguous
4, mixed, ambiguous
7: mixed, ambiguous

Want to focus on 2,4,7 DEGs and see if can glean more information. 

```{r}
DimPlot(
  seurat_integration,
  group.by = "HOXD13_state",
  split.by = "Phase",
  pt.size = 0.25
)

DimPlot(
  seurat_integration,
  group.by = "Phase",
  split.by = "HOXD13_state",
  pt.size = 0.25
)

```

```{r}
DimPlot(
  seurat_integration,
  group.by = "Phase",       # colors = G1 / S / G2M
  split.by = "cell_line",   # panel per cell line
  pt.size = 0.25,
  ncol = 3
) + ggtitle("Cell cycle phase per cell line")

```
*REPETITIVE CODE* just double check everythings right
```{r}
# UMAP with cell-cycle phase and cluster labels

emb = Embeddings(seurat_integration, "umap") %>% 
  as.data.frame()

# consistent axes
emb$umap_1 = emb[,1]
emb$umap_2 = emb[,2]

# add metadata
emb$Phase = seurat_integration$Phase
emb$cluster = seurat_integration$seurat_clusters
emb$cell_line = seurat_integration$cell_line
emb$HOXD13_state = seurat_integration$HOXD13_state

# cluster centers for labelling UMAP

centers = emb %>% 
  group_by(cluster) %>% 
  summarize(umap_1 = mean(umap_1),
            umap_2 = mean(umap_2),
            .groups = "drop")

# plot cell cycle phase w/ cluster labels
ggplot(emb, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(color = Phase), size = 0.25, alpha = 0.8) +
  geom_text(
    data = centers,
    aes(label = cluster),
    color = "black",
    size = 4,
    fontface = "bold"
  ) +
  theme_classic() +
  labs(
    title = "Cell cycle phase with cluster labels",
    color = "Phase"
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)))
```

```{r}

p3 = DimPlot(
  seurat_integration,
  reduction = "umap",
  group.by  = "final_annotation",
  label     = TRUE,
  pt.size   = 0.3
) + ggtitle("Cluster annotations")

p4 = DimPlot(
  seurat_integration,
  reduction = "umap",
  group.by  = "cell_line",
  pt.size   = 0.3
) + ggtitle("EwS cell lines")

p3 | p4
```


*DEGS PER CLUSTER*

Defining by mesenchymal-like or neural-like isn't a very strong label, because of how many genes identified (0-3). It's a good way of having some initial indication, but looking into the top DEGs would be a better indication.


```{r}
# function to get top 5 markers per cluster
top_markers_for_cluster = function(markers, c, n = 5) {
  markers %>% 
    filter(cluster == c) %>% 
    arrange(desc(avg_log2FC)) %>% 
    slice_head(n=n)
}

# use fxn to get 5 DEGS per cluster
top0 = top_markers_for_cluster(up_reg_markers, 0, n=10)
top1 = top_markers_for_cluster(up_reg_markers, 1, n=10)
top2 = top_markers_for_cluster(up_reg_markers, 2, n=10)
top3 = top_markers_for_cluster(up_reg_markers, 3, n=10)
top4 = top_markers_for_cluster(up_reg_markers, 4, n=10)
top5 = top_markers_for_cluster(up_reg_markers, 5, n=10)
top6 = top_markers_for_cluster(up_reg_markers, 6, n=10)
top7 = top_markers_for_cluster(up_reg_markers, 7, n=10)
top8 = top_markers_for_cluster(up_reg_markers, 8, n=10)

head(top0, n=10)
head(top1, n=10)
head(top2, n=10)
head(top3, n=10)
head(top4, n=10)
head(top5, n=10)
head(top6, n=10)
head(top7, n=10)
head(top8, n=10)

top0_genes = c("FBXO32", "MIR34AHG", "TRIM22", "ITGBL1")
top1_genes = c("CAMK4", "TP73", "FAM111B", "E2F2")
top2_genes = c("PIF1", "PSRC1", "ARHGAP11B", "APOLD1")
top3_genes = c("CIB4", "IL1RAPL2", "MROH2A", "HS3ST2")
top4_genes = c("IL3RA", "SYT6", "IER2", "JUNB")
top5_genes = c("SYT4", "PCDH15", "BRINP3", "SLC34A2")
top6_genes = c("PRLR", "CRYBA2", "HCN1", "COL14A1")
top7_genes = c("DMRT2", "UNC13C", "CCDC26", "CA8")
# "identity" markers
top8_genes = c("CENPF", "MKI67", "HIST1H2BN", "HMMR") # proliferation markers, cell-cycle genes

```


```{r}
# final annotation labels
seurat_integration$final_annotation = recode(
  seurat_integration$seurat_clusters,
  "0" = "Mesenchymal-like",
  "1" = "Mesenchymal-like (proliferative)",
  "2" = "Mitotic (proliferative)",
  "3" = "Neural-like",
  "4" = "Stress/mixed program",
  "5" = "Neural-like",
  "6" = "Hybrid neural–mes (proliferative)",
  "7" = "Neural-like",
  "8" = "Highly proliferative"
)

table(seurat_integration$final_annotation)
```

```{r}
# final cluster annotations and ews clusters
p3 = DimPlot(seurat_integration, group.by="final_annotation", pt.size = 0.3, label=TRUE) + ggtitle("Cluster annotations")
p4 = DimPlot(seurat_integration, group.by="cell_line", pt.size = 0.3) + ggtitle("EwS Clusters")

(p3 | p1)

# p2
```
*Looking at some markers for genes to determine expression across clusters*

Are these gene markers unique to clusters or "falsely" being identified as top markers.
Remember top marker expression is relative to other genes w/i same cluster.
```{r}
markers_to_plot = c(
  # cluster 0
  "FBXO32", "ITGBL1",
  # cluster 1
  "CAMK4", "E2F2",
  # cluster 2
  "PIF1",
  # cluster 3
  "IL1RAPL2", "HS3ST2",
  # cluster 4
  "JUNB",
  # cluster 5
  "SYT4", "BRINP3",
  # cluster 6
  "HCN1", "COL14A1",
  # cluster 7
  "DMRT2", "CA8",
  # cluster 8
  "MKI67", "CENPF"
)
```

```{r}
fp = FeaturePlot(
  seurat_integration,
  features = markers_to_plot,
  pt.size = 0.25,
  ncol = 4
) &
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

fp
```

```{r}
vp = VlnPlot(
  seurat_integration,
  features = markers_to_plot,
  group.by = "seurat_clusters",
  pt.size = 0.1,
  ncol = 4
) &
  theme(
    axis.title.x = element_blank()
  )

vp
```
```{r}
FeaturePlot(
  seurat_integration,
  features = c("CENPF", "MKI67", "HIST1H2BN", "HMMR"),
  ncol = 2,
  pt.size = 0.25
)

VlnPlot(
  seurat_integration,
  features = c("CENPF", "MKI67", "HIST1H2BN", "HMMR"),
  group.by = "seurat_clusters",
  ncol = 2,
  pt.size = 0.1
)
```

```{r}
DefaultAssay(seurat_integration)
Reductions(seurat_integration)
```

To further validate/substantiate prev cluster annotations

*MODULE SCORES*

```{r}
# STEP 1: load HOXD13 gene sets
# HOXD13_table = read.csv("/projects/bgmp/shared/groups/2025/sarcoma/tnair/HOXD13_KD_regulated_genes.csv",header=TRUE)
head(HOXD13_table)
names(HOXD13_table) #gene.name, regulation, state

# HOXD13-based EwS programs
hox_neural_genes = HOXD13_table$Gene.name[HOXD13_table$State == "Neural-like"]
hox_mes_genes = HOXD13_table$Gene.name[HOXD13_table$State == "Mesenchymal-like"]

# genes in RNA assay
rna_genes = rownames(seurat_integration[["GeneExpressionMatrix"]])
keep_rna = function(x) intersect(x, rna_genes) # fxn to keep whichever genes intersect b/w gene set + seurat obj

hox_neural_genes = keep_rna(hox_neural_genes)
hox_mes_genes = keep_rna(hox_mes_genes)

length(hox_neural_genes) # how many genes are present in data?
length(hox_mes_genes)
```

```{r}
# STEP 2: MSigDB pathways

# Hallmark gene sets
msig_hallmark = msigdbr(species = "Homo sapiens", category = "H")

hallmark_emt_genes = msig_hallmark %>% filter(gs_name == "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION") %>% pull(gene_symbol)
hallmark_g2m_genes = msig_hallmark %>% filter(gs_name == "HALLMARK_G2M_CHECKPOINT") %>% pull(gene_symbol)
hallmark_e2f_genes = msig_hallmark %>% filter(gs_name == "HALLMARK_E2F_TARGETS") %>% pull(gene_symbol)

# GO: BP (C5) gene sets
# neural programs
msig_go = msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")

go_neurogenesis_genes = msig_go %>% filter(gs_name == "GOBP_NEUROGENESIS") %>% pull(gene_symbol)
go_neuron_diff_genes = msig_go %>% filter(gs_name == "GOBP_NEURON_DIFFERENTIATION") %>% pull(gene_symbol)
go_neuron_dvlpmnt = msig_go %>% filter(gs_name == "GOBP_NEURON_DEVELOPMENT") %>% pull(gene_symbol)
go_neuralcrest = msig_go %>% filter(gs_name == "GOBP_NEURAL_CREST_CELL_FATE_COMMITMENT") %>% pull(gene_symbol)


go_ecm = msig_go %>% filter(gs_name == "GOBP_EXTRACELLULAR_MATRIX_ORGANIZATION") %>% pull(gene_symbol)
go_adhesion = msig_go %>% filter(gs_name == "GOBP_BIOLOGICAL_ADHESION") %>% pull(gene_symbol)
go_cellmig = msig_go %>% filter(gs_name == "GOBP_CELL_MIGRATION") %>% pull(gene_symbol)
go_mesen = msig_go %>% filter(gs_name == "GOBP_MESENCHYME_DEVELOPMENT") %>% pull(gene_symbol)


# STEP 3: Build module scores


# intersect with expressed genes
# addmodulescore can only compute score using genes present in seurat obj
hallmark_emt_genes = keep_rna(hallmark_emt_genes)
hallmark_g2m_genes = keep_rna(hallmark_g2m_genes)
hallmark_e2f_genes = keep_rna(hallmark_e2f_genes)

go_neurogenesis_genes = keep_rna(go_neurogenesis_genes)
go_neuron_diff_genes = keep_rna(go_neuron_diff_genes)
go_neuron_dvlpmnt = keep_rna(go_neuron_dvlpmnt)
go_neuralcrest = keep_rna(go_neuralcrest)

go_ecm = keep_rna(go_ecm)
go_adhesion = keep_rna(go_adhesion)
go_cellmig = keep_rna(go_cellmig)
go_mesen = keep_rna(go_mesen)

# quick check of genes per set
sapply(
  list(
    hox_neural = hox_neural_genes,
    hox_mes = hox_mes_genes,
    H_EMT = hallmark_emt_genes,
    H_G2M = hallmark_g2m_genes,
    H_E2F = hallmark_e2f_genes,
    GO_neurogenesis = go_neurogenesis_genes, # a lot of genes, broad
    GO_neuron_diff = go_neuron_diff_genes, # 0 genes
    GO_neuron_dvlp = go_neuron_dvlpmnt,
    GO_ncrest = go_neuralcrest,
    GO_ecm = go_ecm, # 0 genes
    GO_adh = go_adhesion, # 0 genes
    GO_mesen = go_mesen,
    GO_cellmig = go_cellmig
  ),
  length
)
```

```{r}
# STEP 4: Run AddModuleScore, add to obj

# change default assay to RNA
DefaultAssay(seurat_integration) = "GeneExpressionMatrix"
DefaultAssay(seurat_integration)


# HOXD13 scores
seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(hox_neural_genes),
  name = "HOXD13_neural"
)

seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(hox_mes_genes),
  name = "HOXD13_mes"
)

# Hallmark scores
seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(hallmark_emt_genes),
  name = "H_EMT"
)

seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(hallmark_g2m_genes),
  name = "H_G2M"
)

seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(hallmark_e2f_genes),
  name = "H_E2F"
)

# GO neural/mesenchymal scores

seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(go_neurogenesis_genes),
  name = "GO_neurogenesis"
)

seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(go_neuron_dvlpmnt),
  name = "GO_neuron_dvlpmnt"
)

seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(go_neuralcrest),
  name = "GO_neuralcrest"
)

# mesenchymal-like (?)

seurat_integration = AddModuleScore(
  seurat_integration,
  features = list(go_mesen),
  name = "GO_mesen"
)

tail(colnames(seurat_integration@meta.data))

# seurat adds 1 to the end of the names above, so ex: H_G2M1, H_E2F1
```

```{r}
# STEP 5: visualizations

# HOXD13 neural vs mesenchymal

FeaturePlot(
  seurat_integration,
  features = c("HOXD13_neural1", "HOXD13_mes1"),
  pt.size = 0.3
)


# mesenchymal-ish

FeaturePlot(
  seurat_integration,
  features = c("H_EMT1", "GO_mesen1"),
  pt.size = 0.3
)


# Neural programs
FeaturePlot(
  seurat_integration,
  features = c("GO_neurogenesis1", "GO_neuron_dvlpmnt1", "GO_neuralcrest1"),
  pt.size = 0.3
)

# cell cycle/proliferation
FeaturePlot(
  seurat_integration,
  features = c("H_G2M1", "H_E2F1"),
  pt.size = 0.3
)

```
GO_neurogenesis and GO_neuron_dvlpmnt are TOO broad for cluster assignment

```{r}
DimPlot(seurat_integration, group.by = "seurat_clusters", label = TRUE, repel = TRUE)
```
```{r, fig.height = 12}
VlnPlot(
  seurat_integration,
  features = c("HOXD13_neural1", "HOXD13_mes1", "H_EMT1", "GO_mesen1",
               "GO_neuralcrest1", "H_G2M1", "H_E2F1"),
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3
)

```

```{r, fig.width = 12}
VlnPlot(
  seurat_integration,
  features = c("GO_neurogenesis1", "GO_neuron_dvlpmnt1"),
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3
)
```

```{r}
# define your module score columns
program_cols <- c(
  "HOXD13_neural1", "HOXD13_mes1",
  "H_EMT1", "H_G2M1", "H_E2F1",
  "GO_neurogenesis1", "GO_neuron_dvlpmnt1",
  "GO_neuralcrest1", "GO_mesen1"
)


# compute cluster averages
cluster_program_means <- seurat_integration@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(across(all_of(program_cols), mean, na.rm = TRUE))

# convert to matrix for heatmap
mat <- as.matrix(cluster_program_means[ , -1])
rownames(mat) <- cluster_program_means$seurat_clusters

# scale by row (optional)
pheatmap(mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         border_color = NA,
         fontsize_row = 12,
         fontsize_col = 12)
```

*FINAL ANNOTATIONS*

*VERIFYING MODULE SCORES*
```{r}

# VIEW MODULE SCORES, confusion when looking at gene markers and heatmap, this is more informative
seurat_integration@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(
    HOXD13_neural = median(HOXD13_neural1),
    HOXD13_mes   = median(HOXD13_mes1),
    G2M          = median(H_G2M1),
    E2F          = median(H_E2F1),
    EMT          = median(H_EMT1)
  )

```
```{r}
# FINAL cluster annotations (lineage/program identity)

# recode() transform or replace old values in vector
seurat_integration$final_annotation = 
  recode(seurat_integration$seurat_clusters,
         "0" = "Stress-like",
         "1" = "Mesenchymal-like",
         "2" = "Mesenchymal-like (highly proliferative)",
         "3" = "Neural-like",
         "4" = "Stress-like",
         "5" = "Neural-like",
         "6" = "Mixed neural/mesenchymal",
         "7" = "Neural crest-like",
         "8" = "Highly proliferative")

# order factors so it plots in order, consistent legend order
seurat_integration$final_annotation = factor(
  seurat_integration$final_annotation,
  levels = c(
    "Stress-like",
    "Mesenchymal-like",
    "Mesenchymal-like (highly proliferative)",
    "Neural-like",
    "Mixed neural/mesenchymal",
    "Neural crest-like",
    "Highly proliferative"
  )
)

# check 
table(seurat_integration$seurat_clusters,
      seurat_integration$final_annotation)
# plots

clust_ann = DimPlot(
  seurat_integration,
  group.by = "final_annotation",
  label = TRUE,
  repel = TRUE,
  pt.size = 0.3
) + ggtitle("Cluster annotations")

clust_cls = DimPlot(
  seurat_integration,
  group.by = "cell_line",   # or "orig.ident" / whatever your column is
  label = FALSE,
  pt.size = 0.3
)

clust_ann + clust_cls

```
```{r}
# change color of cell lines for better readability 
cell_colors = c(
  "A4573"  = "#E66101",  # orange
  "A673"   = "#B2182B",  # deep red/magenta
  "CHLA10" = "#1A9850",  # green
  "CHLA9"  = "#6B6DAA",  # slate blue (muted)
  "PDX305" = "#1B9E77",  # teal-green, distinct
  "RDES"   = "#762A83",  # royal purple
  "SKNMC"  = "#66C2A5",  # cyan
  "TC32"   = "#8C510A",  # NEW: deep brown
  "TC71"   = "#FFD92F"   # gold
)

cell_colors2 = c(
  "A4573" = "#E69F00",  # orange
  "A673"  = "#56B4E9",  # light blue
  "CHLA10"= "#009E73",  # bluish green
  "CHLA9" = "#F0E442",  # yellow
  "PDX305"= "#0072B2",  # dark blue
  "RDES"  = "#D55E00",  # reddish orange
  "SKNMC" = "#CC79A7",  # purple/pink
  "TC32"  = "#000000",  # black
  "TC71"  = "#9A6324"   # brown
)

cell_colors3 <- c(
  "A4573"  = "#E66101",  # orange
  "A673"   = "#B2182B",  # red
  "CHLA10" = "#1B7837",  # darker forest green
  "CHLA9"  = "#6B6DAA",  # slate blue
  "PDX305" = "#01665E",  # dark teal
  "RDES"   = "#762A83",  # purple
  "SKNMC"  = "#A6CEE3",  # pale sky blue
  "TC32"   = "#8C510A",  # brown
  "TC71"   = "#FFD92F"   # yellow
)

cell_colors4 <- c(
  "A4573" = "#FF6B6B",   # red
  "A673"  = "#4ECDC4",   # turquoise
  "CHLA10"= "#FFA600",   # orange
  "CHLA9" = "#1A535C",   # deep teal
  "PDX305"= "#9B5DE5",   # purple
  "RDES"  = "#00BBF9",   # cyan
  "SKNMC" = "#FF4D94",   # hot pink
  "TC32"  = "#2EC4B6",   # aqua green
  "TC71"  = "#5F0F40"    # wine-dark purple
)

cl_clusts = DimPlot(
  seurat_integration,
  group.by = "cell_line",
  label = FALSE,
  pt.size = 0.2
) +
  scale_color_manual(values = cell_colors) +
  ggtitle("Cell lines") +
  theme_classic(base_size = 16) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_blank()
  )

clust_ann + cl_clusts
combined_clust_cl = clust_ann + cl_clusts
# save plots
ggsave("final_cluster_annotations.png", plot=clust_ann, width = 13, height = 7, dpi = 300, bg = "white")
ggsave("cl_clusters_final.png", plot=cl_clusts, width = 13, height = 7, dpi = 300, bg = "white")
ggsave("combined_clust_cls.png", plot=combined_clust_cl, width = 15, height = 7, dpi = 300, bg = "white")
```

*save integrated obj for GSEA/GO analysis locally*
```{r}

# check integrated obj has everything i need
Assays(seurat_integration)
Reductions(seurat_integration)
colnames(seurat_integration@meta.data)
grep("final_annotation", colnames(seurat_integration@meta.data), value = TRUE)

head(Idents(seurat_integration))

# save obj (newer than what you transferred to vini)
saveRDS(seurat_integration, file = "seurat_integration_finalAnnotations.rds")
```

*STREAMLINE OBJ FOR LOCAL GSEA/GO USE*

```{r}
# load ANNOTATED seurat integrated obj
seurat_integration = readRDS("seurat_integration_finalAnnotations.rds")
```

```{r}
# RNA only copy
seurat_rna_only = seurat_integration

# remove unneeded assays
# keep RNA assay as default
DefaultAssay(seurat_rna_only) = "GeneExpressionMatrix"
seurat_rna_only[["ATAC"]] = NULL
seurat_rna_only[["MotifMatrix"]] = NULL
seurat_rna_only[["GeneScoreMatrix"]] = NULL
seurat_rna_only[["integratedCCA"]] = NULL

# remove atac related metadata
atac_cols = grep(
  "nDiFrags|nFrags|nMonoFrags|nMultiFrags|FRIP|TSSEnrichment|Peaks|Promoter|Blacklist|ReadsIn",
  colnames(seurat_rna_only@meta.data),
  value = TRUE
)

seurat_rna_only@meta.data[, atac_cols] = NULL

# save slimmed down obj
saveRDS(seurat_rna_only, file="seurat_rna_only_forGSEA_GO.rds")
```

*wanted to try using scCustomize*
```{r}
# using scCustomize to see if plots look cleaner
# clusters 
sc_annot = DimPlot_scCustom(
  seurat_object = seurat_integration,
  group.by      = "final_annotation",
  pt.size       = 0.1,      # adjust if you want denser / lighter points
  label         = TRUE,
  label.size    = 3,
  shuffle       = TRUE,     # shuffle order to reduce overplotting bias
  reduction     = "umap"
) +
  ggtitle("Cluster annotations") +
  Blank_Theme() +                # remove axes/labels
  Move_Legend("right")           # legend on the right

# cell lines
sc_cell = DimPlot_scCustom(
  seurat_object = seurat_integration,
  group.by      = "cell_line",
  pt.size       = 0.1,
  label         = FALSE,
  shuffle       = TRUE,
  reduction     = "umap"
) +
  ggtitle("Cell lines") +
  Blank_Theme() +
  Move_Legend("right")

# side-by-side layout
sc_annot + sc_cell
```

